Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-29 ind_files/02_math_background.tex
Rephrase after proofreading

The second integral in equation \eqref{ito_SDE_integral} is It\^o's `stochastic integral', and is to be interpreted in the following sense: Fix a time $T > 0$. For any $n\geq2$, let $\Pi_n = \{t_1,t_2,\ldots,t_n\}$ be a partition of the interval $[0,T]$. In other words, $\Pi_n$ divides $[0,T]$ into $n$ slices of the form $[t_i,t_{i+1}]$ such that $0 = t_0 < t_1 < t_2 < \ldots < t_n = T$. Then, the (It\^o) stochastic integral from $0$ to $T$ can be thought of as:

@@

That is to say, it is obtained by dividing our time interal into slices of the form $[t_i,t_{i+1}]$, computing the `area of the rectangle' formed with $W_{t_{i+1}}-W_{t_i}$ and $G(t_i,X_{t_i})$ as sides, and then taking the limit\footnote{If you know a little bit of mathematics, it bears noting that this limit is in $L^2(\mathbb{P})$, whereas the corresponding limit in the usual Riemann-Stieltjes integral is evaluated pointwise. If you don't know what this sentence means, just ignore it for the purposes of this thesis :)} of finer and finer slices of time. This should look similar to the classic Riemann integral, with the uniform width $t_{i+1}-t_i$ of the Riemann integral replaced by a random width $W_{t_{i+1}}-W_{t_i}$ corresponding to the (random) displacement of a Brownian particle during the uniform time interval $[t_i,t_{i+1}]$.\\

@@

\subsection{Density-dependence and the intuition for system-size expansions via ecology}\label{sec_intuition_sys_size}
The fundamental idea behind the system-size expansion relates to the nature of the jumps between successive states of a birth-death process. In most situations of evolution, at an individual level, births and deaths of individuals are affected by ecological rules affected by the local population density and not directly by the total population size. Despite this, the jumps themselves occur in terms of the addition (birth) or removal (death) of a \emph{single individual} from the population. If there are many individuals, each individual contributes a negligible amount to the density, and thus, the discontinuous jumps due to individual-level births or deaths can look like a small, \emph{continuous} change in population density. This is the essential idea behind the system-size expansion. The name derives from the formalization of this idea as a change of variable from the discrete values $\{0,1,2,\ldots,n-1,n,n+1,\ldots\}$ to the approximately continuous values $\{0,1/K,2/K,\dots,x-1/K,x,x+1/K,\ldots\}$ by the introduction of a `system size parameter' $K$. In physics and chemistry, $K$ is usually the total volume of a container in which physical or chemical reactions take place and is thus a `hard' limit. In ecology, we will only impose a `soft' limit by requiring that births and deaths must scale with $K$ in a way that the population almost surely cannot grow indefinitely if $K$ is finite, reflecting the empirical fact that the total amount of resources in the world is limited (of course, this should also be reflected in how birth and death rates vary as functions of population density). Thus, in our systems, $K$ will manifest as some fundamental limit on resources, such as habitat size or carrying capacity. When $K$ is large, the fact that transitions occur in units of a small value $1/K$ can be exploited via a Taylor expansion of the transition rates in the Master equation, which then yields a Fokker-Planck equation upon neglecting higher order terms\footnote{If this sounds handwavy to you, see chapter 11, section 3 in~\citet{ethier_markov_1986} for a more rigorous treatment.}. A similar approximation is well-known (ever since Fisher) in theoretical population genetics, where it goes by the name of the `diffusion approximation'\citep{ewens_mathematical_2004} or `continuum limit'~\citep{czuppon_understanding_2021}, and has been heavily used by Kimura~\citep{crow_introduction_1970} in his stochastic models. However, the population genetics version of the approximation usually either relies on total population size being fixed~\citep{crow_introduction_1970, lande_natural_1976,ewens_mathematical_2004} or is conducted in an ad-hoc manner without specifying an explicit system size parameter.

@@

Consider a population of identical individuals subject to some ecological rules that affect individuals' birth and death rates. I will neglect any potential factor that could lead to two individuals becoming non-identical, like mutations, etc. Since all individuals are identical, we only really need to track the total population size through time to know everything there is to know about the population. The population as a whole at any time $t$ can thus be characterized by a single number - its population size (Figure \ref{fig_1D_pop_description}). Imagine further that if a population has $n$ identical individuals, then, from the ecological rules, we can determine a \emph{birth rate} $b(n)$, which gives us a measure of the probability that a new individual will be born and the population size becomes $n+1$ `per unit time'. One must be slightly precise about what exactly they mean when they say `per unit time' since there are no discrete `time steps' for individuals to be born. Here, by `birth rate', I mean the probability that there will be a birth (and no death) per an \emph{infinitesimal} amount of time. More formally, letting $N_t$ denote the random variable representing the population size at time $t$ and letting $\mathbb{P}(E)$ denote the probability (in the common-sense usage) of an event $E$, the birth rate\footnote{Note that unlike usual ecology convention, this is \emph{not} a per-capita birth rate. The ecology per capita birth and death rates can be found by dividing my birth/death rates by the current population size $n(t)$} $b(n)$ of individuals a population with population size $n$ is the quantity
\begin{equation}
\label{1D_birthrate_defn}
b(n) \coloneqq \lim_{\epsilon \to 0}\frac{1}{\epsilon}\mathbb{P}\left(N_{t+\epsilon}=n+1 | N_{t}=n\right)
\end{equation}
Exactly analogously, we can also define the \emph{death rate} $d(n)$ of individuals in a population of $n$ individuals as the quantity
\begin{equation}
\label{1D_deathrate_defn}
d(n) \coloneqq \lim_{\epsilon \to 0}\frac{1}{\epsilon}\mathbb{P}\left(N_{t+\epsilon}=n-1 | N_{t}=n\right)
\end{equation}
An alternative, perhaps more intuitive characterization, of these same quantities is the following: If we have a population of size $n$, and we know that \emph{either a birth or a death} has just occurred, then, the probability that the event that occurred is a birth is
\begin{equation*}
\mathbb{P\big[\textrm{ birth } \big{|} \textrm{ something happened }\big]} = \frac{b(n)}{b(n)+d(n)}
\end{equation*}
and the probability that the event was instead a death is given by
\begin{equation*}
\mathbb{P\big[\textrm{ death } \big{|} \textrm{ something happened }\big]} = \frac{d(n)}{b(n)+d(n)}
\end{equation*}
\begin{example}\label{ex_1D_stoch_logistic}
	Consider the case where the per-capita birth rate is a constant $\lambda > 0$, \emph{i.e}, $b(n) = \lambda n$, and the per-capita death rate has the linear density-dependence $d(n) = \left(\mu + (\lambda-\mu)\frac{n}{K}\right)n$, where $\mu$ and $K$ are positive constants. Taking the difference between the birth and death rates, we obtain $b(n) - d(n) = (\lambda - \mu)n\left(1-\frac{n}{K}\right)$, where, identifying $r=\lambda-\mu$, we obtain the familiar logistic equation on the RHS. Note, however, that the population itself is stochastic, whereas the logistic equation is a deterministic description.
\end{example}
Now, let $P(n,t)$ be the probability that the population size is $n$ at time $t$. We wish to have an equation to describe how $P(n,t)$ changes with time - this will provide a probabilistic description of how we expect the population size to change over time.

\myfig{0.97}{figures/2.1_BD_process_1D.png}{\textbf{Schematic description of a one-dimensional birth-death process}. Consider a population of identical individuals. The state of the system can be described by a single number, the population size (numbers within the circles). Births and deaths result in changes in the total population size, and the birth and death rates (arrows) are dependent on the current population size. For a given state $n$, the \textcolor{blue}{blue} arrows depict the rate of `inflow' to the state (from the blue states), whereas the \textcolor{red}{red} arrows depict the rate of `outflow'.}{fig_1D_pop_description}
To do this, we imagine a large ensemble of populations. In a large ensemble of copies evolving independently, a fraction $P(n,t)$ will have population size $n$ at time $t$ by definition of probability. We can now simply measure the `inflow' and `outflow' of copies of the population from each state~\ref{fig_1D_pop_description}. If a population has $n$ individuals, it could either have gotten there from a population of $n+1$ individuals, with a death rate of $d(n+1)$, or from a population of $n-1$ individuals, with a birth rate of $b(n-1)$. Thus, the rate of `inflow' to the state $n$ is given by
\begin{equation}
\label{1D_rate_in}
R_{\textrm{in}}(n,t) = b(n-1)P(n-1,t) + d(n+1)P(n+1,t)
\end{equation}
Similarly, if the population has $n$ individuals, it could obtain a different state in two ways: With rate $b(n)$, the population witnesses a birth, and with rate $d(n)$, it witnesses a death. Thus, the rate of `outflow' is given by
\begin{equation}
\label{1D_rate_out}
R_{\textrm{out}}(n,t) = b(n)P(n,t) + d(n)P(n,t)
\end{equation}
The rate of change of the probability of the system being in state $n$ is given by the rate of inflow minus the rate of outflow. Thus, we have
\begin{align}
\frac{\partial P}{\partial t}(n,t) &= R_{\textrm{in}}(n,t) - R_{\textrm{out}}(n,t)\nonumber\\
&= b(n-1)P(n-1,t) + d(n+1)P(n+1,t) - b(n)P(n,t) - d(n)P(n,t)\label{1D_M_eqn_nostep}
\end{align}
For convenience, let us define two `step operators' $\mathcal{E}^{\pm}$, which act on any functions of populations to their right by either adding or removing an individual, \textit{i.e}
\begin{equation*}
\mathcal{E}^{\pm}f(n,t) = f(n \pm 1,t)
\end{equation*}
Rearranging the RHS of \eqref{1D_M_eqn_nostep} to write in terms of these step operators, we obtain the compact expression
\begin{equation}
\label{1D_M_eqn}
\frac{\partial P}{\partial t}(n,t) = (\mathcal{E}^{-}-1)b(n)P(n,t) + (\mathcal{E}^{+}-1)d(n)P(n,t)
\end{equation}
This is the so-called `master equation', and completely describes our system. However, in general, $b(n)$ and $d(n)$ may be rather complicated, in which case it may not be possible to solve \eqref{1D_M_eqn} directly.

>>>>

The second integral in equation \eqref{ito_SDE_integral} is It\^o's `stochastic integral', and is to be interpreted in the following sense: Fix a time $T > 0$. For any $n\geq2$, let $\Pi_n = \{t_1,t_2,\ldots,t_n\}$ be a partition of the interval $[0,T]$. In other words, the points contained in $\Pi_n$ divide $[0,T]$ into $n$ slices of the form $[t_i,t_{i+1}]$ such that $0 = t_0 < t_1 < t_2 < \ldots < t_n = T$. Then, the (It\^o) stochastic integral of the function $G(t,x)$ over the time interval $[0,T]$ is given by

@@

That is to say, it is obtained by dividing our time interval into slices of the form $[t_i,t_{i+1}]$, computing the `area of the rectangle' formed with $W_{t_{i+1}}-W_{t_i}$ and $G(t_i,X_{t_i})$ as sides, and then taking the limit\footnote{If you are familiar with some real analysis, it bears noting that this limit is in $L^2(\mathbb{P})$, whereas the corresponding limit in the usual Riemann-Stieltjes integral is evaluated pointwise. If you don't know what this sentence means, just ignore it for the purposes of this thesis :)} of finer and finer slices of time. This should look similar to the classic Riemann integral, with the uniform width $t_{i+1}-t_i$ of the Riemann integral replaced by a random width $W_{t_{i+1}}-W_{t_i}$ corresponding to the (random) displacement of a Brownian particle during the uniform time interval $[t_i,t_{i+1}]$.\\

@@

\subsection{Density-dependence and the intuition for system-size expansions via ecology}\label{sec_intuition_sys_size}
The fundamental idea behind the system-size expansion relates to the nature of the jumps between successive states of a birth-death process. In most situations of interest to population dynamics, at an individual level, births and deaths of individuals are affected by ecological rules that depend on the local population density and not directly on the total population size. Despite this, the jumps themselves occur in terms of the addition (birth) or removal (death) of a \emph{single individual} from the population. If there are many individuals, each individual contributes a negligible amount to the density, and thus, the discontinuous jumps due to individual-level births or deaths can look like a small, \emph{continuous} change in population density. This is the essential idea behind the system-size expansion. The name derives from the formalization of this idea as a change of variable from the discrete values $\{0,1,2,\ldots,n-1,n,n+1,\ldots\}$ to the approximately continuous values $\{0,1/K,2/K,\dots,x-1/K,x,x+1/K,\ldots\}$ through the introduction of a `system size parameter' $K$. In physics and chemistry, $K$ is usually the total volume of a container in which physical or chemical reactions take place and is thus a `hard' limit on the number of discrete values allowed. In ecology, we will only impose a `soft' limit by requiring that births and deaths must scale with $K$ in a way that the population almost surely cannot grow indefinitely if $K$ is finite, reflecting the empirical fact that the total amount of resources in the world is limited (of course, this should also be reflected in how birth and death rates vary as functions of population density). Thus, in our systems, $K$ will manifest as some fundamental limit on resources, such as habitat size or carrying capacity. When $K$ is large, the fact that transitions occur in units of a small value $1/K$ can be exploited via a Taylor expansion of the transition rates in the Master equation, which then yields a Fokker-Planck equation upon neglecting higher order terms\footnote{If this sounds handwavy to you, see chapter 11, section 3 in~\citet{ethier_markov_1986} for a more rigorous treatment.}. A similar approximation is well-known (ever since Fisher) in theoretical population genetics, where it goes by the name of the `diffusion approximation'~\citep{ewens_mathematical_2004} or `continuum limit'~\citep{czuppon_understanding_2021}, and has been heavily used by Kimura~\citep{crow_introduction_1970} in his stochastic models. However, the population genetics version of the approximation usually either relies on total population size being fixed~\citep{crow_introduction_1970, lande_natural_1976,ewens_mathematical_2004} or is conducted in an ad-hoc manner without specifying an explicit system size parameter.

@@

Consider a population of identical individuals subject to some ecological rules that affect individuals' birth and death rates. I will neglect any potential factor that could lead to two individuals becoming non-identical, like mutations, etc. Since all individuals are identical, we only really need to track the total population size through time to know everything there is to know about the population. The population as a whole at any time $t$ can thus be characterized by a single number - its population size (Figure \ref{fig_1D_pop_description}). Imagine further that if a population has $n$ identical individuals, then, from the ecological rules, we can determine a \emph{birth rate} $b(n)$, which gives us a measure of the probability that a new individual will be born and the population size becomes $n+1$ `per unit time'. One must be slightly precise about what exactly they mean when they say `per unit time' since there are no discrete `time steps' for individuals to be born. Here, by `birth rate', I mean the probability that there will be a birth (and no death) per an \emph{infinitesimal} amount of time. More formally, letting $N_t$ denote the random variable representing the population size at time $t$ and letting $\mathbb{P}(E)$ denote the probability (in the common-sense usage) of an event $E$, the birth rate\footnote{Note that unlike usual ecology convention, this is \emph{not} a per-capita birth rate. The ecology per capita birth and death rates can be found by dividing my birth/death rates by the current population size $n(t)$} $b(n)$ of individuals a population with population size $n$ is the quantity
\begin{equation}
\label{1D_birthrate_defn}
b(n) \coloneqq \lim_{\epsilon \to 0}\frac{1}{\epsilon}\mathbb{P}\left(N_{t+\epsilon}=n+1 | N_{t}=n\right)
\end{equation}
Exactly analogously, we can also define the \emph{death rate} $d(n)$ of individuals in a population of $n$ individuals as the quantity
\begin{equation}
\label{1D_deathrate_defn}
d(n) \coloneqq \lim_{\epsilon \to 0}\frac{1}{\epsilon}\mathbb{P}\left(N_{t+\epsilon}=n-1 | N_{t}=n\right)
\end{equation}
An alternative, perhaps more intuitive characterization, of these same quantities is the following: If we have a population of size $n$, and we know that \emph{either a birth or a death} is about to occur, then, the probability that the event that occurs is a birth is
\begin{equation*}
\mathbb{P\big[\textrm{ birth } \big{|} \textrm{ something happened }\big]} = \frac{b(n)}{b(n)+d(n)}
\end{equation*}
and the probability that the event is instead a death is
\begin{equation*}
\mathbb{P\big[\textrm{ death } \big{|} \textrm{ something happened }\big]} = \frac{d(n)}{b(n)+d(n)}
\end{equation*}
\begin{example}\label{ex_1D_stoch_logistic}
	Consider the case where the per-capita birth rate is a constant $\lambda > 0$, \emph{i.e}, $b(n) = \lambda n$, and the per-capita death rate has the linear density-dependence $\left(\mu + (\lambda-\mu)\frac{n}{K}\right)$, \emph{i.e.} the total death rate is $d(n) = \left(\mu + (\lambda-\mu)\frac{n}{K}\right)n$, where $\mu$ and $K$ are positive constants. Taking the difference between the birth and death rates, we obtain $b(n) - d(n) = (\lambda - \mu)n\left(1-\frac{n}{K}\right)$, where, identifying $r=\lambda-\mu$, we obtain the familiar logistic equation on the RHS. Note, however, that the population itself is stochastic, whereas the logistic equation is a deterministic description.
\end{example}
Now, let $P(n,t)$ be the probability that the population size is $n$ at time $t$. We wish to have an equation to describe how $P(n,t)$ changes with time - this will provide a probabilistic description of how we expect the population size to change over time.

\myfig{0.97}{figures/2.1_BD_process_1D.png}{\textbf{Schematic description of a one-dimensional birth-death process}. Consider a population of identical individuals. The state of the system can be described by a single number, the population size (numbers within the circles). Births and deaths result in changes in the total population size, and the birth and death rates (arrows) are dependent on the current population size. For a given state $n$, the \textcolor{blue}{blue} arrows depict the rate of `inflow' to the state (from the blue states), whereas the \textcolor{red}{red} arrows depict the rate of `outflow'.}{fig_1D_pop_description}
To do this, we imagine a large ensemble of populations. In a large ensemble of copies evolving independently, a fraction $P(n,t)$ will have population size $n$ at time $t$ by definition of probability. We can now simply measure the `inflow' and `outflow' of copies of the population from each state (Figure~\ref{fig_1D_pop_description}). If a population has $n$ individuals, it could either have gotten there from a population of $n+1$ individuals, with a death rate of $d(n+1)$, or from a population of $n-1$ individuals, with a birth rate of $b(n-1)$. Thus, the rate of `inflow' to the state $n$ is given by
\begin{equation}
\label{1D_rate_in}
R_{\textrm{in}}(n,t) = b(n-1)P(n-1,t) + d(n+1)P(n+1,t)
\end{equation}
Similarly, if the population has $n$ individuals, it could obtain a different state in two ways: With rate $b(n)$, the population witnesses a birth, and with rate $d(n)$, it witnesses a death. Thus, the rate of `outflow' is given by
\begin{equation}
\label{1D_rate_out}
R_{\textrm{out}}(n,t) = b(n)P(n,t) + d(n)P(n,t)
\end{equation}
The rate of change of the probability of the system being in state $n$ is given by the rate of inflow minus the rate of outflow. Thus, we have
\begin{align}
\frac{\partial P}{\partial t}(n,t) &= R_{\textrm{in}}(n,t) - R_{\textrm{out}}(n,t)\nonumber\\
&= b(n-1)P(n-1,t) + d(n+1)P(n+1,t) - b(n)P(n,t) - d(n)P(n,t)\label{1D_M_eqn_nostep}
\end{align}
For convenience, let us define two `step operators' $\mathcal{E}^{\pm}$, which act on any functions of populations to their right by either adding or removing an individual, \textit{i.e}
\begin{equation*}
\mathcal{E}^{\pm}f(n,t) = f(n \pm 1,t)
\end{equation*}
Rearranging the RHS of \eqref{1D_M_eqn_nostep} to write in terms of these step operators, we obtain the compact expression
\begin{equation}
\label{1D_M_eqn}
\frac{\partial P}{\partial t}(n,t) = (\mathcal{E}^{-}-1)b(n)P(n,t) + (\mathcal{E}^{+}-1)d(n)P(n,t)
\end{equation}
This is the so-called `master equation', and completely describes our system. However, in general, $b(n)$ and $d(n)$ may be rather complicated, in which case it may not be possible to solve \eqref{1D_M_eqn} directly.
